Confronting preferential sampling when analyzing population distributions: diagnosis and model-based triage \footnotemark[2]

### 2017-04-24 manuscript/ReviewerResponses.tex

%  make sure that the document has 25 lines per page (it is 12 pt)

\setlength{\textheight}{575pt} \setlength{\baselineskip}{23pt}

%  IMPORTANT -- place the \title, \author, etc, \begin{summary}...
%  \end{summary} \keywords statements in the order demonstrated below
\rm
Here, we provide line by line responses to associate editor and reviewer comments, suggestions, and edits.  For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).  In addition, our revision includes copy editing corrections from our editorial office and some clarifications in response to additional comments we received from an interested reader. We provide a marked up PDF version of our manuscript showing all changes made.

In the process of conducting revisions we corrected an error in our estimation code (somewhere in the model development and testing process, the link function for $R_i$ was misspecified as multinomial-logit instead of logit).  Correcting this error has led to several qualitative changes to the paper.  Conclusions from the simulation study were strengthened; however, bearded seal conclusions were made more murky.  In particular, both of the model types incorporating preferential sampling (linear and trend surface) appeared to produce biologically unrealistic estimates in the bearded seal example.  This was particularly troubling because AIC favored these models.  We decided to conduct an additional cross validation analysis to quantify predictive ability, and found that non-preferential models produced more accurate predictions.  We now argue against using preferential sampling models for the case study.

@@

[From the comments to author section]
Your paper definitely focuses on an important issue, and provides ways to address the problem. Besides the issues raised by the referee, I think you should relate your paper a bit more to the ecological literature, in particular on SDMs (see Edwards et al. 2006). I think in particular that in addition to the types of sampling you discuss, one could also consider sampling designs that are optimal to estimate response curve (eg Albert et al. 2010 - sorry for the self-citation...). Ecological studies may have different objectives, and I think you should try to make your paper relevant to as many readers as possible.

Albert, C. H., N. G. Yoccoz, T. C. Edwards, C. H. Graham, N. E. Zimmermann, and W. Thuiller. 2010. Sampling in ecology and evolution – bridging the gap between theory and practice. Ecography 33:1028-1037.
Edwards, T. C., D. R. Cutler, N. E. Zimmermann, L. Geiser, and G. G. Moisen. 2006. Effects of sample survey design on the accuracy of classification tree models in species distribution models. Ecological Modelling 199:132-141.

\textcolor{blue}{We thank the editor for pointing us to these articles - our knowledge of the species distribution modeling literature is admittedly a little lacking.  These papers are certainly germane to the present study, and we have included additional material in the introduction (2nd paragraph) that talks about (1) implications of design vs. model-based inference, and (2) the importance of the particular objective function (e.g., minimizing variance, estimating species-habitat relationships) for optimal design.}

\textcolor{blue}{
Additionally, we have made the following changes to try to make the manuscript accessible to a wider audience:
\begin{itemize}
  \item An additional figure (fig. 1) to more clearly depict preferential sampling for visual learners.
  \item  In the methods section, we motivate inclusion of spatially autocorrelated random effects by indicating ``Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates."
  \item We changed ``it is usually impossible to view system state perfectly even in locations where sampling is conducted" to ``it is usually impossible to conduct a complete census of all animals present in a surveyed location"
  \item We provide examples of survey and observer specific covariates. (visibility, observer experience)
  \item We changed the ``underlying spatial field of interest" and ``spatially-referenced random field values" to ``the process of interest"
  \item We changed ``surveyed locations were selected independently of the abundance generating process" to ``surveyed locations were randomly selected (i.e. independent of abundance)."
  \item We changed ``In this case, the abundance process and sampling intensity submodels were independent, as is the with case canonical SDMs (at least when fitted to presence-absence or count data)" to ``In this case, sampling intensity is assumed to be independent of underlying abundance, as implicitly assumed by most SDMs in the ecological literature."
  \item Numerous other small changes to wording are documented in the track changes document.
\end{itemize}}

@@

\textcolor{blue}{There are several points to make here.  First, we never claimed that we were fitting the exact model proposed by \citet{RoyleBerliner1999} - we're not.  However, the covariance structure implied by the two model \emph{is} actually very similar.  For instance, if $\bfeta \sim \text{Gaussian}({\bf 0},\bfSigma_\eta)$ and $\bfdelta \sim \text{Gaussian}({\bf 0},\bfSigma_\delta)$, then
\begin{equation}
  \text{Cov}\left(
    \begin{array}{c}
    \bfeta + \bfB \bfdelta \\
    \bfdelta
    \end{array}
   \right)
   =
   \left(
   \begin{array}{cc}
      \bfSigma_\eta + \bfB \bfSigma_\delta \bfB^\prime & \bfB \bfSigma_\delta \\
         (\bfB \bfSigma_\delta)^\prime & \bfSigma_\delta
    \end{array}
   \right),
\end{equation}
as desired.  Second, it seems less important that the model be exactly the same as it does to provide a motivation (and citation) for how the $\bfB$ matrix can be specified.  We think both of these points were adequately accounted for in our initial submission, as we (a) described the \citet{RoyleBerliner1999} model as being ``similar" to the one we proposed here, and (b) used their paper as a reference for how the $\bfB$ matrix can be formulated and interpreted.
}

In your application, line 199, you report difficulties fitting a model with even modest structure on $\bfB$. Might a
joint model for $\bfeta$ and $\bfdelta$, as specified by \citet{RoyleBerliner1999}, allow you to specify a more complicated
model for $\bfB$?

\textcolor{blue}{
  As suggested above, we believe our model to be very similar to that proposed by \citet{RoyleBerliner1999}.  We're not seeing how drawing the random effects jointly would necessarily improve performance (we're also having trouble seeing how we could implement this in our chosen software).  Also, after further investigation of the \citet{RoyleBerliner1999} paper, we're not sure they ever actually fit a model with a more complicated relationship in $\bfB$.  In fact, their one example on ozone-temperature analysis uses the same simple diagonal parameterization we used here.  We've adjusted text in several places to reflect this. For instance, we state that ``However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not been investigated" when talking about higher order preferential sampling models.
}

\textcolor{blue}{
  After correcting the error in our estimation model link function (see below), qualitative results have changed somewhat in the bearded seal example, with both the simple single parameter $b$ model and the trend surface model producing similar, vexing results. So, the question of identifiability may actually be one of estimability - i.e., how often do these methods produced reasonable estimates for real world datasets?  We know the single parameter $b$ model works $\approx 80\%$ of the time with simulated data; were we simply unlucky with the bearded seal data set? Or do other factors (e.g., overdispersion, line transects vs. point counts) affect the quality of inference?  In the discussion, we now suggest that identifiability and reliability of inference of such models be investigated in future research.
}

@@

On line 139, you suggest “we often need to fix $\beta_0^* = 0.0$ . . . to permit parameter identification”. The word
``often" suggests you can sometimes include the intercept parameter when estimating random effects. Can
you provide more guidance about when you would need to fix the intercept at 0, and how a practitioner
would know they need to fit the intercept at 0?

\textcolor{blue}{
  Initially we were trying to fit ICAR type models which have a known parameter redundancy built in, such that the intercept is confounded with the spatial process \citep{RueHeld2005}.  However, the type of Gaussian process model we settled on and used in analysis does not suffer from this deficiency, but confusion resulted because somewhere along the way we mistakenly built in a multinomial logit link function into the estimation code for the $R_i$ (whose intercept is also not identifiable).  We have rectified this mistake in the present version, and all models for $R_i$ are now Bernoulli with a logit link, so that the intercept term is now identifiable.  We have rerun all analyses with this corrected configuration.
}

@@

\textcolor{blue}{
  We have replaced the Poisson observation model with a binomial one in eqn 4 to make this clearer, and have indicated dependence on $Z_i$ through conditioning.  The reason that we had the Poisson in there originally is that the models we fit to data use the fact that the marginal distribution for $Y_i$ given the Poisson-Binomial mixture is itself a Poisson distribution.  However, we think it will be clearer to present the model hierarchically at this stage and then to describe our marginalization (something many ecologists may not care about) when describing software.  We have created a new ``Software" methods subsection which describes this in greater detail. Importantly, we describe how predictions are made; when detection probability is equal to 1.0, predictions are made using the formulation $\hat{N}_i = Y_i + (1-A_i)\exp(\hat{\mu}_i)$.  This framework incorporates a finite population correction and is similar to that employed by \citet{JohnsonEtAl2010} in the context of point process likelihood analysis of distance sampling data.  It also prevents predicted abundance in a cell from being less than the observed count.
}

On line 61, you indicate $\bfmu$ and $\bfnu$ represent log-scale abundance (associated with {\bf Z}) and the logit of the probability of sampling (associated with {\bf R}). For equation 5, however, you use $\bfmu_i$ to represent the mean value of a Gaussian process, which created some confusion. You might consider more generally allowing $\bfmu$ to be a function of covariates influencing the state variable of interest, and $\bfnu$ to be a function of covariates
influencing placement of sampling locations. You could then replace $\psi_i$ in equation 6 with $\nu_i$, which would make it immediately clear you were discussing factors influencing the probability of site placement.

\textcolor{blue}{
  We appreciate the reviewer's attempt to help us streamline notation and to make things clearer for the reader.  We have implemented the basic premise of this idea, although some of the particulars differ from the reviewer's suggestion.  Specifically we now describe $\bfmu$ and $\bfnu$ as indicating ``the link-scale process of interest and the logit of the probability of sampling, respectively.  As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects."  Then in the next section we use $\mu_i$ in eq. 5 and $\nu_i$ in eq. 6, although we retain $\xi_i$, since in the next section we also wish to write $\nu_i$ in terms of shared spatial effects.  This is all a little tricky, in that the \citet{DiggleEtAl2010} paper considers a continuous support model (whereas ours is discrete), and the structure is slightly different.  Admittedly, maintaining recognizable notation while keeping precise definitions is a challenge.
}

@@

Line 65: This sentence seems to discount the value of good study design, which is still important for model-
based estimation.

\textcolor{blue}{
We've altered text to be a little less cavalier, now stating that
``Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement."
}

Line 77: It's unclear to me why the distinction between fine and coarse-scale preferential sampling is important. They both seem to represent the process of dependence between sampling locations and the process
of interest.

\textcolor{blue}{
As the size of the sampling unit decreases (with the limit being a point process model in continuous space) there is indeed little difference.  But practically, it seems likely that one could use a reasonable design to select sampling units, but still open themselves to bias if they target non-representative habitat within individual sampling units.  However, this point is well taken, and we don't think the distinction necessarily needs the prominence we've given it.  We now address this topic for the first time in the discussion.
}

@@

Line 195: You only describe 4 models below, not 6.

\textcolor{blue}{We now report results form all 6 models.}

Simulation study results: can you report bias for other parameters (e.g. slope parameters) in your model,
even if this is relegated to an appendix?

\textcolor{blue}{Yes, we have created an additional appendix table (S3) that summarizes bias in $b$ (the preferential sampling parameter), as well as bias in the species-habitat parameters.  We also provide convergence statistics in this appendix. Thanks for this suggestion; in particular, it helps show that preferential sampling is not necessarily a big deal if one is only interested in drawing inference on parameters informing species-habitat relationships.  %We note that the bias in the slope parameter of the species-habitat relationship %is likely because of the well-known confounding that occurs between regression %coefficients and spatially autocorrelated random effects in spatial models %\citep[e.g.][]{ReichEtAl2006,HodgesReich2010}.
}

@@

Paragraph starting on line 219: I find this paragraph very confusing. In particular, it is unclear to me exactly
what you mean when you state you ``...refitted $M_{cov=1;b=0}$ to produce an estimate of apparent abundance
without this feature. Specifically, we refitted the model with 20 pseudo-absences in this portion of the study
area to better inform abundance-covariate relationships". Does Fig. 5 reflect this re-fitted model? Why not
fit all models in this manner for consistency?

\textcolor{blue}{Our initial thought was that inserting ``pseudo-absences" could possibly impact inference about the preferential sampling parameter because those locations were not actually sampled and shouldn't affect the $R_i$ model. So, we conducted model selection without pseudo-absences, but then inserted them when making predictions. In the revision, we've circumvented this issue by introducing depth as an additional covariate (with linear and quadratic effects). This seems to have remedied the issue with positive estimates of abundance in offshore areas.  So, no need to mess around with pseudo-absences anymore.}

@@

Line 225: a model-averaged estimate of abundance?

\textcolor{blue}{After correcting the link function in present model runs, it no longer makes sense to present a model-averaged estimate.}

Table 1: I find these tables are easier to read if models are listed in decreasing order of $\Delta$AIC.

\textcolor{blue}{We now present models in order of decreasing predictive ability (as measured with cross-validated mean square prediction error)}

@@

Fig. 3: In my version, the blue and green dashed lines do not show up in the legend.

\textcolor{blue}{Strange, this is not the case in our version.  Guessing it is an issue with PDF compilation on the MEE submission website.  We'll work with copy editors to ensure it looks right if it heads into production.}

### 2017-04-24 manuscript/AppendixS1_revision.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

$^1$ National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA

@@

In this appendix, we provide further details on model development, implementation, and simulation testing for the preferential sampling model outlined in the main article text. Our modeling approach works by constructing (a) a species distribution model (SDM) describing abundance or presence/absence that includes spatially autocorrelated random effects, and (b) a model for the probability of site selection.  Both submodels can be expressed in terms of habitat or landscape covariates.  However, site selection probability can also be written as a function of the same spatially autocorrelated random effects in (a).  Including the same set of random effects in both submodels is the mechanism for allowing dependence betwen the two models, and allows us to estimate and control for the potentially biasing effects of preferential sampling.

@@

We focus here on a discrete spatial domain; extensions to point process models in continuous space should be relatively straightforward.  In particular, we suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  Let $Z_i$ denote the state variable of interest in cell $i$ (e.g., occupancy, abundance), and let $Y_i$ denote observations made in cell $i$.  In some cases, observations may be equivalent to the state variable in sampled cells (i.e. when detection probability is 1.0).  In others, an observation model $[Y_i | Z_i, \boldsymbol{\theta}]$ relating $Y_i$ to the underlying process of interest will be required, where $\boldsymbol{\theta}$ are nuisance parameters describing detection.  For instance, for survey counts in our simulation study and in the bearded seal example, we set
\begin{eqnarray*}
  [Y_i | Z_i, \boldsymbol{\theta}] & = & \textrm{Binomial}(Z_i, A_i p_i),
\end{eqnarray*}
where $A_i$ gives the proportion of survey unit $i$ that is sampled and $p_i$ is a detection probability ($p_i$ is set to 1.0 for the simulation study and an informative prior distribution is specified in the bearded seal example).

@@

\hspace{0.5in} We model the site selection probabilities in the same manner.  First, we write $R_i$, as
\begin{eqnarray}
 \label{eq:R}
  [R_i|\nu_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
Here, $\beta_0^*$ is an intercept, ${\bf x}_i^*$ is a vector of covariates (in practice, these can be the same or different from ${\bf x}_i$), and $\boldsymbol{\eta}=\{ \eta_1,\eta_2,\hdots,\eta_S \}$ are spatially autocorrelated random effects.  The connection between the two models (eqs. \ref{eq:mu} and \ref{eq:nu}) is the reliance of both on $\delta_i$.  In eq. \ref{eq:nu}, the matrix ${\bf B}$ describes this relationship.  Previous preferential sampling models in the statistical literature have set $\textbf{B} = b \textbf{I}$, where $\textbf{I}$ is an identity matrix and $b$ is a parameter to be estimated.  This is the approach we take in simulation study and in several of the models fit to real data.  Another approach is to construct $\textbf{B}$ so it changes smoothly in space, as in \citep{RoyleBerliner1999}.  In this manner, the degree of preferential sampling could vary across the landscape; presumably, one would need fairly rich data to be able to model a spatially varying relationship.

\hspace{0.5in} We have yet to specify a model for spatial autocorrelation.  For analyses in this paper, we model mean-zero spatial random effects (i.e., the $\delta_i$ and $\eta_i$) for each site $i$ as multivariate normal:
\begin{eqnarray*}
  \boldsymbol{\delta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\delta}), \text{ and} \\
  \boldsymbol{\eta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\eta}),
\end{eqnarray*}
where $\boldsymbol{\Sigma}_{\lambda}$ and $\boldsymbol{\Sigma}_{\eta}$ are covariance matrices.  As is typical of spatial models in statistics, we allow the magnitude of spatial autocorrelation to decrease as a function of the distance between survey units.  In particular, we use the well known Mat\'{e}rn covariance function to model covariance as a function of distance.  In addition to the distances between survey units (which we calculate relative to survey unit centroids), the Mat\'{e}rn covariance function uses a range (or scale) parameter $\kappa$, a precision parameter $\tau$ (inverse marginal variance), and a smoothness parameter to model covariance.  For analyses in this paper, we set the smoothness parameter to 1.0 and treat $\kappa$ and $\tau$ as parameters to be estimated.  To improve computational efficiency, we use a Gaussian Markov random field (GMRF) representation of the underlying Gaussian field \citep{LindgrenEtAl2011}, as described by \citet{ThorsonEtAl2015} (Appendix B).

\hspace{0.5in} To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}. Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  Requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

@@

\begin{enumerate}
  \item Generate a spatially autocorrelated covariate over a $25 \times 25$ grid, where the covariate at grid cell $i$, $X_i$, is generated via a multivariate normal distribution:
      \begin{eqnarray}
        \textbf{X} & \sim & \textrm{MVN} (\textbf{0},\boldsymbol{\Sigma}).
        \label{eq:cov}
      \end{eqnarray}
      An isotropic, Gaussian covariance structure is used for $\boldsymbol{\Sigma}$, where the $i$th row and $j$th column of $\boldsymbol{\Sigma}$ is set to $\Sigma_{ij}=\sigma^2 \exp{-(d_{ij}/s)^2}$, where $d_{ij}$ is the distance between the centroid of grid cell $i$ and grid cell $j$, $\sigma$ is a standard deviation (here, set to 1.0), and $s$ is a scale parameter (here, set to 5.0).
  \item Generate true abundance in each survey unit (grid cell), $N_i$, as a function of the covariate from step 1, together with additional spatially autocorrelated error.  We generate a vector of such abundances as
      \begin{eqnarray}
        N_i & \sim & \textrm{Poisson}(\exp(\beta_0 + X_i \beta_1 + \delta_i)),
      \end{eqnarray}
      where the $\boldsymbol{\delta}$ are drawn in an identical manner to $\textbf{X}$ in step 1.  We set $\beta_0 = 2.0$ and drew $\beta_1 \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation.
  \item For each simulation, we selected 50 grid cells for sampling (i.e., set $R_i = 1.0$) according to 3 alternative survey configurations.  Each survey configuration, grid cells were drawn randomly without replacement, where the probability of sampling cell $i$ is proportional to
      \begin{equation*}
        \exp(X_i \beta_1^*  + \eta_i + b \delta_i).
      \end{equation*}
      Here, $\eta_i$ is a spatially autocorrelated random effect drawn in an identical manner to $\textbf{X}$ in step 1, and $b$ describes the level of preferential sampling.  Three levels are used: $b=0$, which makes the sampling model conditionally independent from the abundance process (i.e., no preferential sampling); $b=1$, which induces some dependence between the abundance and sampling process (i.e., moderate preferential sampling); $b=5$ which makes cells of high abundance much more likely to be sampled than low abundance cells (i.e., pathological preferential sampling).   We drew $\beta_1^* \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation replicate.  An illustration of the consequences of different survey configurations is provided in Fig. 2 of the main manuscript.
  \item For each of the 3 sampling configurations, we generated a count dataset.  In each case, we generated a count $Y_i$ for each cell selected for sampling (i.e., for cells where $R_i=1.0$) according to
      \begin{eqnarray*}
        Y_i & \sim & \textrm{Binomial}(N_i,A).
      \end{eqnarray*}
      Here, $A$ represents the portion of the grid cell that is surveyed, and was set to 0.5 for all simulations and grid cells.
  \item For each count data set, we fitted 2 different estimation models in TMB (see \textit{Preferential sampling model development}).  In the first, the preferential sampling parameter $b$ was fixed to 0.0 (thus in effect assuming no preferential sampling).  In the second, $b$ was estimated.
  \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  This bias correction is also used to obtain abundance estimates in the bearded seal example
\end{enumerate}

>>>>

$^1$ Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, NOAAA, National Marine Fisheries Service, 2725 Montlake Boulevard E, Seattle, WA 98112, USA

@@

In this appendix, we provide further details on model development, implementation, and simulation testing for the preferential sampling model outlined in the main article text. Our modeling approach works by constructing (a) a species distribution model (SDM) that explains variation in survey counts, and (b) a model for the probability of site selection.  Both submodels can be expressed in terms of habitat or landscape covariates and spatially autocorrelated random effects.  Including the same set of random effects in both submodels is the mechanism for allowing dependence betwen the two models, and allows us to estimate and control for the potentially biasing effects of preferential sampling.

@@

We focus here on a discrete spatial domain; extensions to point process models in continuous space should be relatively straightforward.  In particular, we suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  Let $Z_i$ denote the state variable of interest in cell $i$ (e.g., occupancy, abundance), and let $Y_i$ denote observations made in cell $i$.  In some cases, observations may be equivalent to the state variable in sampled cells (i.e. with complete area coverage and detection probability equal to 1.0).  In others, an observation model $[Y_i | Z_i, \boldsymbol{\theta}]$ relating $Y_i$ to the underlying process of interest will be required, where $\boldsymbol{\theta}$ are nuisance parameters describing detection.  For instance, for survey counts in our simulation study and in the bearded seal example, we set
\begin{eqnarray*}
  [Y_i | Z_i, \boldsymbol{\theta}] & = & \textrm{Binomial}(Z_i, A_i p_i),
\end{eqnarray*}
where $A_i$ gives the proportion of survey unit $i$ that is sampled and $p_i$ is a detection probability.  Note that for ease of exposition we simply set $p_i = 1.0$ for analyses in this paper, but this constraint should be relaxed in the future if such estimates are to be used in conservation or management.

@@

\hspace{0.5in} We model the site selection probabilities in the same manner.  First, we write $R_i$, as
\begin{eqnarray}
 \label{eq:R}
  [R_i|\nu_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
Here, $\beta_0^*$ is an intercept, ${\bf x}_i^*$ is a vector of covariates (in practice, these can be the same or different from ${\bf x}_i$), and $\boldsymbol{\eta}=\{ \eta_1,\eta_2,\hdots,\eta_S \}$ are spatially autocorrelated random effects.  The connection between the two models (eqns. \ref{eq:mu} and \ref{eq:nu}) is the reliance of both on $\delta_i$.  In eqn. \ref{eq:nu}, the matrix ${\bf B}$ describes this relationship.  Previous preferential sampling models in the statistical literature have set $\textbf{B} = b \textbf{I}$, where $\textbf{I}$ is an identity matrix and $b$ is a parameter to be estimated.  This is the approach we take in our simulation study and in several of the models fit to real data.  Another approach is to construct $\textbf{B}$ so it changes smoothly in space, as suggested (but apparently not implemented) by \citep{RoyleBerliner1999}.  In this manner, the degree of preferential sampling could vary across the landscape; presumably, one would need fairly rich data to be able to model a spatially varying relationship.

\hspace{0.5in} We have yet to specify a model for spatial autocorrelation.  For analyses in this paper, we model mean-zero spatial random effects (i.e., the $\delta_i$ and $\eta_i$) for each site $i$ as multivariate normal:
\begin{eqnarray*}
  \boldsymbol{\delta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\delta}), \text{ and} \\
  \boldsymbol{\eta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\eta}),
\end{eqnarray*}
where $\boldsymbol{\Sigma}_{\lambda}$ and $\boldsymbol{\Sigma}_{\eta}$ are covariance matrices.  As is typical of spatial models in statistics, we allow the magnitude of spatial autocorrelation to decrease as a function of the distance between survey units.  In particular, we use the well known Mat\'{e}rn covariance function to model covariance as a function of distance.  In addition to the distances between survey units (which we calculate relative to survey unit centroids), the Mat\'{e}rn covariance function uses a range (or scale) parameter $\kappa$, a precision parameter $\tau$ (inverse marginal variance), and a smoothness parameter to model covariance.  For analyses in this paper, we set the smoothness parameter to 1.0 and treat $\kappa$ and $\tau$ as parameters to be estimated.  To improve computational efficiency, we use a Gaussian Markov random field (GMRF) representation of the underlying Gaussian field \citep{LindgrenEtAl2011}, as described by \citet{ThorsonEtAl2015} (Appendix B).  For increased stability we estimated only one $\kappa$ value (i.e. the same $\kappa$ parameter was used for $[{\bf Z}]$ and $[{\bf R}]$.

\hspace{0.5in} To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}. Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  Requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

@@

\begin{enumerate}
  \item Generate a spatially autocorrelated covariate over a $25 \times 25$ grid, where the covariate at grid cell $i$, $X_i$, is generated via a multivariate normal distribution:
      \begin{eqnarray}
        [\textbf{X}] & = & \textrm{MVN} (\textbf{0},\boldsymbol{\Sigma}).
        \label{eq:cov}
      \end{eqnarray}
      An isotropic, Gaussian covariance structure is used for $\boldsymbol{\Sigma}$, where the $i$th row and $j$th column of $\boldsymbol{\Sigma}$ is set to $\Sigma_{ij}=\sigma^2 \exp{-(d_{ij}/s)^2}$, where $d_{ij}$ is the distance between the centroid of grid cell $i$ and grid cell $j$, $\sigma$ is a standard deviation (here, set to 1.0), and $s$ is a scale parameter (here, set to 5.0).
  \item Generate true abundance in each survey unit (grid cell), $N_i$, as a function of the covariate from step 1, together with additional spatially autocorrelated error.  We generate a vector of such abundances from the distributions
      \begin{eqnarray}
        [N_i] & = & \textrm{Poisson}(\exp(\beta_0 + X_i \beta_1 + \delta_i)),
      \end{eqnarray}
      where the $\boldsymbol{\delta}$ are drawn in an identical manner to $\textbf{X}$ in step 1.  We set $\beta_0 = 2.0$ and drew $\beta_1$ from a $\textrm{Uniform}(-0.5,0.5)$ distribution for each simulation.
  \item For each simulation, we selected 50 grid cells for sampling (i.e., set $R_i = 1.0$) according to 3 alternative survey configurations.  Each survey configuration, grid cells were drawn randomly without replacement, where the probability of sampling cell $i$ is proportional to
      \begin{equation*}
        \exp(X_i \beta_1^*  + \eta_i + b \delta_i).
      \end{equation*}
      Here, $\eta_i$ is a spatially autocorrelated random effect drawn in an identical manner to $\textbf{X}$ in step 1, and $b$ describes the level of preferential sampling.  Three levels are used: $b=0$, which makes the sampling model conditionally independent from the abundance process (i.e., no preferential sampling); $b=1$, which induces some dependence between the abundance and sampling process (i.e., moderate preferential sampling); $b=3$ which makes cells of high abundance much more likely to be sampled than low abundance cells (i.e., pathological preferential sampling).   We drew $\beta_1^*$ from a  $\textrm{Uniform}(-0.5,0.5)$ distribution for each simulation replicate.  An illustration of the consequences of different survey configurations is provided in Fig. 2 of the main manuscript.
  \item For each of the 3 sampling configurations, we generated a count dataset.  In each case, we generated a count $Y_i$ for each cell selected for sampling (i.e., for cells where $R_i=1.0$) from the distribution
      \begin{eqnarray*}
        [Y_i|N_i] & = & \textrm{Binomial}(N_i,A).
      \end{eqnarray*}
      Here, $A$ represents the portion of the grid cell that is surveyed, and was set to 0.5 for all simulations and grid cells.
  \item For each count data set, we fitted 2 different estimation models in TMB (see \textit{Preferential sampling model development}).  In the first, the the models for $[{\bf R}]$ and $[{\bf N}]$ were configured to be independent.  In the second, a single parameter $b$ was estimated, where ${\bf B} = b {\bf I}$ in eqn. (\ref{eq:nu}).
  \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  This bias correction is also used to obtain abundance estimates in the bearded seal example.  Estimates of $N$ are made relative to $N_i = Y_i + (1-A) \exp(\beta_0 + X_i \beta_1 + \delta_i)$, which includes an implicit finite population correction \citep{JohnsonEtAl2010}.
\end{enumerate}
